---
output:
  pdf_document: default
  html_document: default
---
---
title: "Study of Applicants Applying for Masters and Phd programs at U of R"
output:
  pdf_document: 
    latex_engine: xelatex
  html_document:
    df_print: paged
subtitle: 
fontsize: 9 pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, eval=T,tidy=TRUE, tidy.opts=list(width.cutoff=60))
```
Veronica Mata Ramirez

```   {r}
library(tidyverse)  
library(readr)      
library(ggplot2)    
library(dplyr)

data <- read.csv("application_data_2015_2022.csv")

# Summary of the dataset
#summary(data)
str(data)

# Number of samples
cat("Number of total samples:", nrow(data), "\n")

masters <- data %>% filter(Degree == "Master's")
phd <- data %>% filter(Degree == "PhD")
cat("Number of Master's records:", nrow(masters), "\n")
cat("Number of PhD records:", nrow(phd), "\n")


print("Count of missing values by column") 
empty_or_na_counts <- sapply(data, function(x) sum(x == "" | is.na(x)))
sorted_counts <- sort(empty_or_na_counts, decreasing = TRUE)
print(sorted_counts)

# Function to create braplot
create_bar_plot <- function(data, column_name, top_n = 15, fill_column = "Entry.Term") {
  # Check if the column names exist in the data frame
  if (!column_name %in% names(data)) {
    stop("The specified column does not exist in the data frame.")
  }
  if (!fill_column %in% names(data)) {
    stop("The specified fill column does not exist in the data frame.")
  }
  
  prepared_data <- data %>%
    filter(!is.na(!!sym(column_name)) & !!sym(column_name) != "") %>%
    group_by(!!sym(column_name), !!sym(fill_column)) %>%
    summarise(Count = n(), .groups = "drop") %>%
    arrange(desc(Count)) %>%
    top_n(top_n, wt = Count) # Select top N based on Count

  ggplot(prepared_data, aes(x = reorder(!!sym(column_name), Count), y = Count, fill = !!sym(fill_column))) +
    geom_bar(stat = "identity", position = position_dodge(), width = 0.9) +
    theme_minimal() +
    labs(title = paste("Top", top_n, "Categories in", column_name, "by", fill_column), 
         x = column_name, 
         y = "Count") +
    coord_flip() +
    theme(axis.text.y = element_text(size = 10, angle = 45))
}


create_bar_plot(data, "Birth.Country", 30,  "Degree")
create_bar_plot(data, "Sex", 4,  "Degree")
create_bar_plot(data, "Program..ASE.", 50,  "Entry.Term")
create_bar_plot(data, "Institution.1.Name", 50,  "Entry.Term")


# Function to see distributions
create_filtered_histogram <- function(data, plot_column, filter_column, filter_value, binwidth = 0.5, fill_color = "pink") {
  # Filter the data
  filtered_data <- data %>%
    filter(!!sym(filter_column) == filter_value)
  
  # Create the histogram
  ggplot(filtered_data, aes(x = !!sym(plot_column))) +
    geom_histogram(binwidth = binwidth, fill = fill_color, color = "black") +
    theme_minimal() +
    labs(title = paste("Distribution of", plot_column, "for", filter_column, "=", filter_value), 
         x = plot_column, 
         y = "Frequency") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}


# To help respond 2nd sub question of th primary question
create_filtered_histogram(data, "Age.at.App.Submission", "Degree", "Master's", 0.5, "pink")
create_filtered_histogram(data, "Age.at.App.Submission", "Degree", "PhD", 0.5, "blue")

# To help respond 3rd sub question of th primary question
create_filtered_histogram(data, "Age.at.App.Submission", "Have.you.ever.failed.a.course.", 0, 0.5, "green")
create_filtered_histogram(data, "Age.at.App.Submission", "Have.you.ever.failed.a.course.", 1, 0.5, "yellow")


ggplot(data, aes(x = Degree, y = Age.at.App.Submission)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Age Distribution by Degree", 
       x = "Degree", 
       y = "Age at Application Submission") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))



ggplot(data, aes(x = Citizenship, y = Age.at.App.Submission)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Age Distribution by Citizenship", 
       x = "Citizenship", 
       y = "Age at Application Submission") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```









```{r}
library(DescTools)
library(dplyr)
data <- read.csv("D:\\StatsProject\\StatisticsProject\\ProjectforFall2023Dataset20220816204620.csv")

sapply(data,class)

#Print the names of columns in the dataset
column_names <- colnames(data)
print(column_names)

# Missing values 
print("Count of missing values by column") 
empty_or_na_counts <- sapply(data, function(x) sum(x == "" | is.na(x)))
sorted_counts <- sort(empty_or_na_counts, decreasing = TRUE)
print(sorted_counts)

#Removing some columns 
#Column to Remove
column_to_remove <- c("Ref","Citizenship2","Type","How.Applicant.Heard...Other","Institution.3.Degree","Institution.3.Major","Institution.3.Location","Institution.3.Level.of.Study","Institution.3.Name","Institution.2.Degree","Institution.2.Location","Institution.2.Major","Institution.2.Level.of.Study","Institution.2.Name"," Other.Schools.Applying.To","Job.3.Title","Other.Schools.Applying.To","Job.2.Title","Job.1.Title ","Job.1.Title","Institution.1.Major","Institution.1.Level.of.Study","Native.Language","Spouse.Studying.Applying","Any.Relatives.Listed.")
data <- data[, !colnames(data) %in% column_to_remove]
column_names <- colnames(data)
print(column_names)


```
```{r}
#Replacing the missing values in the age column (Mean)
mean_age <- mean(data$Age.at.App.Submission , na.rm = TRUE)
rounded_mean_age <- round(mean_age)
cat("The rounded mean age is:", rounded_mean_age)

# Replacing missing age values with the rounded value 
data$Age.at.App.Submission <- ifelse(is.na(data$Age.at.App.Submission), rounded_mean_age, data$Age.at.App.Submission)

# check code 
print("Count of missing values by column") 
empty_or_na_counts <- sapply(data, function(x) sum(x == "" | is.na(x)))
sorted_counts <- sort(empty_or_na_counts, decreasing = TRUE)


# Function to find mode
calculate_mode <- function(x) {
  tbl <- table(x)
  mode_value <- as.character(names(tbl[tbl == max(tbl)]))
  return(mode_value)
}

# Find mode of the "Sex" column
mode_sex <- calculate_mode(data$Sex)

# Print the result
cat("Mode of the 'Sex' column:", mode_sex)

# Replace empty places with the mode value
data$Sex[data$Sex == ""] <- mode_sex

# check code 
print("Count of missing values by column") 
empty_or_na_counts <- sapply(data, function(x) sum(x == "" | is.na(x)))
sorted_counts <- sort(empty_or_na_counts, decreasing = TRUE)
```

```{r}
#Replacing missing values in Have you ever failed a course column
# Function to find mode
calculate_mode <- function(x) {
  tbl <- table(x, useNA = "ifany")
  mode_value <- as.numeric(names(tbl[tbl == max(tbl)]))
  return(mode_value)
}

# Find mode of the column
mode_value <- calculate_mode(data$Have.you.ever.failed.a.course.)

# Replace missing values with the mode
data$Have.you.ever.failed.a.course.[is.na(data$Have.you.ever.failed.a.course.)] <- mode_value

# check code 
print("Count of missing values by column") 
empty_or_na_counts <- sapply(data, function(x) sum(x == "" | is.na(x)))
sorted_counts <- sort(empty_or_na_counts, decreasing = TRUE)
```

```{r}
#Replacing missing values in Ever.Placed.on.Academic.Probation 
# Function to find mode
calculate_mode <- function(x) {
  tbl <- table(x, useNA = "ifany")
  mode_value <- as.numeric(names(tbl[tbl == max(tbl)]))
  return(mode_value)
}

# Find mode of the column
mode_value <- calculate_mode(data$Ever.Placed.on.Academic.Probation )

# Replace missing values with the mode
data$Ever.Placed.on.Academic.Probation [is.na(data$Ever.Placed.on.Academic.Probation )] <- mode_value

# check code 
print("Count of missing values by column") 
empty_or_na_counts <- sapply(data, function(x) sum(x == "" | is.na(x)))
sorted_counts <- sort(empty_or_na_counts, decreasing = TRUE)


frequency_table <- table(data$How.Applicant.Heard.About.UR)

# Find the entry with the maximum frequency
most_frequent_entry <- names(frequency_table)[which.max(frequency_table)]

# Print the result
cat("The most frequent entry is:", most_frequent_entry)
cat("With a frequency of:", frequency_table[most_frequent_entry])
```

```{r}
# Use the table function to get the frequency of each unique entry
frequency_table <- table(data$How.Applicant.Heard.About.UR)

# Create a data frame to display the results
result_df <- data.frame(
  Entry = names(frequency_table),
  Frequency = as.vector(frequency_table)
)

# Print the result data frame
print(result_df)

#Replace missing values (including empty strings) with "unknown sources"
data$How.Applicant.Heard.About.UR <- ifelse(data$How.Applicant.Heard.About.UR == "" | is.na(data$How.Applicant.Heard.About.UR), "unknown sources", data$How.Applicant.Heard.About.UR)

#Check the count of missing values by column
print("Count of missing values by column") 
empty_or_na_counts <- sapply(data, function(x) sum(x == "" | is.na(x)))
sorted_counts <- sort(empty_or_na_counts, decreasing = TRUE)

#Replacing the Time Status missing values
unique(data$Time.Status)
missing_values <- c("", "NA")
data$Time.Status[data$Time.Status %in% missing_values] <- mode_value

#check code 
print("Count of missing values by column") 
empty_or_na_counts <- sapply(data, function(x) sum(x == "" | is.na(x)))
sorted_counts <- sort(empty_or_na_counts, decreasing = TRUE)

#Replacing missing values in Degree
data$Degree[is.na(data$Degree) | data$Degree == ""] <- names(sort(table(data$Degree), decreasing = TRUE)[1])

#Check missing values by column
print("Count of missing values by column:")
empty_or_na_counts <- sapply(data, function(x) sum(x == "" | is.na(x)))
sorted_counts <- sort(empty_or_na_counts, decreasing = TRUE)
```

```{r}
#Replacing missing values in Institution 1 name 
data$Institution.1.Name  <- ifelse(data$Institution.1.Name  == "" | is.na(data$Institution.1.Name ), "unknown sources", data$Institution.1.Name )

# check code 
print("Count of missing values by column") 
empty_or_na_counts <- sapply(data, function(x) sum(x == "" | is.na(x)))
sorted_counts <- sort(empty_or_na_counts, decreasing = TRUE)

# Renaming 'citizenship' to 'Citizenship_status'
data <- data %>%
  rename(Citizenship_status =  Citizenship )

# Renaming 'Citizenship1' to 'citizenship'
data <- data %>%
  rename(citizenship = Citizenship1)

# Print the updated column names
colnames(data)


```
```{r}
# Replace 'birth_country' missing values 
data$Birth.Country[is.na(data$Birth.Country)] <- data$Birth.Country[is.na(data$Birth.Country)]

# check code 
print("Count of missing values by column") 
empty_or_na_counts <- sapply(data, function(x) sum(x == "" | is.na(x)))
sorted_counts <- sort(empty_or_na_counts, decreasing = TRUE)

#Replace 'Sub.Category' missing values 
data$Sub.Category  <- ifelse(data$Sub.Category  == "" | is.na(data$Sub.Category ), "Not Specified", data$Sub.Category )

# check code 
print("Count of missing values by column") 
empty_or_na_counts <- sapply(data, function(x) sum(x == "" | is.na(x)))
sorted_counts <- sort(empty_or_na_counts, decreasing = TRUE)
```

```{r}
#Handling missing values in Recommender feature 
library(dplyr)

data <- data %>%
  mutate(Num_Two_Recs = (Recommender.1.Relationship != "" & Recommender.2.Relationship != "") +
                        (Recommender.1.Relationship != "" & Recommender.3.Relationship != "") +
                        (Recommender.2.Relationship != "" & Recommender.3.Relationship != ""),
         Num_Three_Recs = (Recommender.1.Relationship != "" & Recommender.2.Relationship != "" & Recommender.3.Relationship != ""))

# Print the updated data frame
print(data)

# check code 
print("Count of missing values by column") 
empty_or_na_counts <- sapply(data, function(x) sum(x == "" | is.na(x)))
sorted_counts <- sort(empty_or_na_counts, decreasing = TRUE)

#dropping columns 
column_to_remove <- c("Recommender.1.Relationship","Recommender.2.Relationship","Recommender.3.Relationship")
data <- data[, !colnames(data) %in% column_to_remove]
column_names <- colnames(data)
print(column_names)


```
```{r}
# Combine values and handle missing values 
# this column is a combination of birth and citizenship both columns
# two diffeent countries side by side pedict that birth country and citizenship is different for that person
data$combined <- ifelse(is.na(data$Birth.Country), data$citizenship,
                      ifelse(is.na(data$citizenship), data$Birth.Country,
                             ifelse(data$Birth.Country == data$citizenship, data$Birth.Country,
                                    paste(data$Birth.Country, data$citizenship, sep = "/"))))

# Print the result
print(data)

# Specify the column names to remove (can omit if required)
columns_to_remove <- c("Birth.Country", "citizenship")

# Remove the specified columns
data <- data[, !(names(data) %in% columns_to_remove)]

# Print the final column names
cat("\nFinal Column Names:\n")
print(names(data))


# check code 
print("Count of missing values by column") 
empty_or_na_counts <- sapply(data, function(x) sum(x == "" | is.na(x)))
sorted_counts <- sort(empty_or_na_counts, decreasing = TRUE)
```
```{r}
#Handling missing values in citizenship_status
unique_values <- unique(data$Citizenship_status )
print(unique_values)
frequency_table <- table(data$Citizenship_status )
print(frequency_table)
data$Citizenship_status   <- ifelse(data$Citizenship_status   == "" | is.na(data$Citizenship_status  ), "FN", data$Citizenship_status  )

# check code 
print("Count of missing values by column") 
empty_or_na_counts <- sapply(data, function(x) sum(x == "" | is.na(x)))
sorted_counts <- sort(empty_or_na_counts, decreasing = TRUE)

# Handling missing values in combined
data$combined   <- ifelse(data$combined   == "" | is.na(data$combined  ), "Unknown", data$combined  )

# check code 
print("Count of missing values by column") 
empty_or_na_counts <- sapply(data, function(x) sum(x == "" | is.na(x)))
sorted_counts <- sort(empty_or_na_counts, decreasing = TRUE)

# Handling missing values in Institution.1.Location 
data$Institution.1.Location   <- ifelse(data$Institution.1.Location   == "" | is.na(data$Institution.1.Location  ), "Not Mentioned", data$Institution.1.Location  )

# check code 
print("Count of missing values by column") 
empty_or_na_counts <- sapply(data, function(x) sum(x == "" | is.na(x)))
sorted_counts <- sort(empty_or_na_counts, decreasing = TRUE)

# Handling missing values in Institution.1.Degree
data$Institution.1.Degree  <- ifelse(data$Institution.1.Degree  == "" | is.na(data$Institution.1.Degree  ), "Not Specified", data$Institution.1.Degree)

# check code 
print("Count of missing values by column") 
empty_or_na_counts <- sapply(data, function(x) sum(x == "" | is.na(x)))
sorted_counts <- sort(empty_or_na_counts, decreasing = TRUE)

#Handling missing values in Num_Three_Recs  and Num_Two_Recs
data <- data[complete.cases(data$Num_Two_Recs), ]
data <- data[complete.cases(data$Num_Three_Recs), ]

# check code 
print("Count of missing values by column") 
empty_or_na_counts <- sapply(data, function(x) sum(x == "" | is.na(x)))
sorted_counts <- sort(empty_or_na_counts, decreasing = TRUE)

# Handling missing values in Decision.1
data <- data[complete.cases(data$Decision.1) & data$Decision.1 != "", ]

# Check the count of missing values by column
print("Count of missing values by column") 
empty_or_na_counts <- sapply(data, function(x) sum(x == "" | is.na(x)))
sorted_counts <- sort(empty_or_na_counts, decreasing = TRUE)
```

```{r}
AgeVariance <- data.frame(data$Degree,data$Age.at.App.Submission)
AgeVariance


# Summary statistics
summary(AgeVariance)

# Boxplot to visualize age variance by degree type
boxplot(data$Age.at.App.Submission ~ data$Degree , data = data, col = c("lightblue", "lightgreen","Violet"), main = "Age Variance by Degree Type", xlab = "Degree Type", ylab = "Age")

# t-test to compare age variance between Master's and PhD groups
anova_result <- aov(data$Age ~ data$Degree, data = AgeVariance)
print(summary(anova_result))
```
```{r}
scatter_plot <- ggplot(AgeVariance, aes(x = data$Age.at.App.Submission, y = data$Degree, color = data$Degree)) +
  geom_point() +
  labs(title = "Scatter Plot of Age at Application Submission by Degree",
       x = "Age at Application Submission",
       y = "Degree") +
  theme_minimal()

scatter_plot
```

```{r}
# Assuming "data" is your data frame
cross_tab <- table(data$Sex, data$Degree)
chi_squared_result <- chisq.test(cross_tab)

# Display the cross-tabulation
print(cross_tab)

# Display the chi-squared test result
print(chi_squared_result)

```
```{r}
# Assuming "data" is your data frame
correlation_age_recommendation <- cor(data[, c("Age.at.App.Submission", "Num_Three_Recs")])

# Display the correlation matrix
print(correlation_age_recommendation)


correlation_age_recommendation <- cor(data[, c("Age.at.App.Submission", "Num_Two_Recs")])
print(correlation_age_recommendation)

```

```{r}
# Fit a multiple linear regression model
regression_model <- lm(data$Age.at.App.Submission ~ data$Degree + data$Have.you.ever.failed.a.course. + data$Program..ASE. + data$Sex + data$Entry.Term, data = data)

# Display the summary of the regression model
summary(regression_model)

```


The regression results provide information about the relationship between the dependent variable (Age.at.App.Submission) and various independent variables. Here are some key points to interpret the results:

Coefficients:
Intercept (32.03): The estimated average age at application submission when all other variables are zero.

Degree:

Master's: For every one-unit increase in holding a Master's degree, the average age decreases by approximately 9.29 units.

MS: The effect is a decrease of approximately 5.98 units.

PhD: The effect is a decrease of approximately 7.05 units.

Have_You_Ever_Failed (0.87): For every one-unit increase in the variable "Have_You_Ever_Failed," the average age increases by 0.87 units.

Program_ASE:

Data Science: The effect is an increase of approximately 0.44 units.

Electrical and Computer Engineering: The effect is a decrease of approximately 0.63 units.

Sex (0.21): Males have an average age that is approximately 0.21 units higher than females.

Entry_Term:
Various categories: The coefficients represent the estimated change in average age compared to the reference category.


Statistical Significance:
Significance is indicated by the p-values associated with each coefficient.
Variables with small p-values (typically < 0.05) are considered statistically significant.
In the results, most variables are statistically significant (indicated by the asterisks).

Model Fit:
Multiple R-squared: Represents the proportion of the variance in the dependent variable explained by the model. In this case, around 13.15% of the variance in age is explained.

Adjusted R-squared: Takes into account the number of predictors in the model. Adjusted R-squared is slightly lower.

F-statistic: Tests the overall significance of the model. A small p-value (< 0.05) indicates that at least one variable in the model is 

significant. Here, the p-value is < 2.2e-16, suggesting overall significance.

Residuals:
Residuals represent the differences between observed and predicted values.






