---
title: "Study Of Applicants Applying for Masters and PhD programs at UofR"
author: "Anik De, Babli Dey, Sharon Gilbert, Veronica Mata"
date: "2023-12-09"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- importing student application data from the csv file "application_data_2015_2022.csv" -->

```{r importing student dataset, echo=FALSE}
applicantData<-read.csv("application_data_2015_2022.csv")
```

<!-- importing all the library here -->
```{r import require library, echo=FALSE}
library(ggplot2)
```

## Visualization of the data before beginning any analysis (will include unknown data) - Masters v/s PhD:


1. What is the annual total no. of applicants, which term recorded the highest and lowest number of applicants?
```{r no.Of applicants, echo=FALSE}

```


2. Identify the top 10 undergraduate institutes from which applicants are applying.
```{r top 10 undergrad institutes, echo=FALSE}

```


3. How do the numbers of applicants vary by gender across different terms?
```{r applicants by gender, echo=FALSE}

```


4. What are the top 10 countries from which applicants have applied to the university?
```{r  top country of applicants, echo=FALSE}

```


5. Determine the total number of applicants who were placed on probation.
```{r  applicants placed on probation, echo=FALSE}

```


6. Determine the total number of applicants who failed a course.
```{r  applicants failed a course, echo=FALSE}

```

## Data Cleaning: Comment on the steps and also mention if you have modified the data set


1. Apply a filter to the dataset based on the 24 specified dimensions.
```{r  filter dimensions, echo=FALSE}

```


2. Change column names to lowercase and use underscores or simpler terms for better interpretability.
```{r  change the column names, echo=FALSE}

```


3. Segment the dataset into two parts: one for master's programs and another for Ph.D. programs.
```{r  segment dataset into masters and phd program, echo=FALSE}

```
  
  
4. For empty values, categorize them as "Unknown" for categorical data. For age, fill in with the average age corresponding to the    respective program.
```{r  deal with null values, echo=FALSE}

```

## Descriptive Analysis: Master's and PhD


1. Calculate the mean, variance, and standard deviation of age, and visualize the distribution using histograms and box plots.
```{r  descriptive analysis of age, echo=FALSE}

```


2. Examine the acceptance rate based on gender across different terms.
```{r  acceptance rate by gender, echo=FALSE}

```


3. Explore the acceptance rate by country in relation to different terms.
```{r  acceptance rate by country, echo=FALSE}

```
  
  
4. Investigate the acceptance rate for applicants who were placed on probation.
```{r  acceptance rate of probation applicants, echo=FALSE}

```


5. Investigate the acceptance rate for applicants who failed a course.
```{r  acceptance rate applicants who failed a course, echo=FALSE}

```


6. Investigate the age distribution between applicants who failed a course v/s applicants who did not fail a course.
```{r  age distribution failed vs not failed a course, echo=FALSE}

```

## Inferential Analysis:

1. Is there a notable distinction in age variance between applicants who have experienced course failure and those who haven't, considering each degree type?
```{r  age variance hypothesis test, echo=FALSE}

```


2. Is there a discernible difference in the average age between master's and Ph.D. students?
```{r  age variance hypothesis test by program, echo=FALSE}

```


3. Compare the acceptance rate for students who learned about the university through family/friends versus those who found it through an internet search.
```{r  hypothesis test based on how they hear about U of R, echo=FALSE}

```

## Logistic Regression:

1. Can we use factors such as age, citizenship, gender, recommender relationships, current employment status, and program choice to predict the likelihood of a candidate being accepted?
```{r  logistic regression, echo=FALSE}
#create more chunks as required
```













  






